---
output: github_document
bibliography: inst/REFERENCES.bib
csl: inst/apa-old-doi-prefix.csl
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sdmTools
#### Author: João Frederico Berner

<!-- badges: start -->
[![R-CMD-check](https://github.com/jfberner/jfbpack/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jfberner/jfbpack/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of sdmTools is to allow easy implementation of methods not implemented in the sdm package, using the standard outputs of the sdm:: functions. The methods are all described in the literature, and are usually not compiled in any single package, which leads to very long coding and implementing sessions. The author felt ENM analyses and mapping currently sparsed in the literature could be centralized in a single package.

There are four currently working "branches/tools" in the package. Namely, 

 1. **Boyce Index** [[@hirzelEvaluatingAbilityHabitat2006]](https://doi.org/10.1016/j.ecolmodel.2006.05.017) , with the function `sdm_to_boyceIndex()`;
 2. **Consistency/Congruence Maps** [[@morales-barberoInputMattersMatter2019a]](https://doi.org/10.1111/2041-210X.13124) with the function  `sdm_to_consistency()`. [Original source code here](https://besjournals.onlinelibrary.wiley.com/action/downloadSupplement?doi=10.1111%2F2041-210X.13124&file=mee313124-sup-0002-AppendixB_Rcode.R). I intend to allow new congruence maps to be created from a set of environmental predictors with a new function in the near future, but for now only the [original congruence map](https://doi.org/10.5061/dryad.6kv7k29) is available; (authorization pending)
 3. **Accumulation of Occurrences Curve** [[@jimenezLeavingAreaReceiving2020a]](https://doi.org/10.1111/2041-210X.13479) with the functions `sdm_to_occ.pnts()` and `sdm_to_output.mods()`, to be passed onto the [original functions](https://github.com/LauraJim/SDM-hyperTest) `accum.occ()`, and later `comp.accplot()`. (authorization pending)
 4. **Frequency Ensemble** [[@sobral-souzaSpeciesExtinctionRisk2015]](https://doi.org/10.1016/j.ncon.2015.11.009) with functions `sdm_to_freqEnsemble()` and `frequency_ensemble_plot()`. The approach is the same as in the paper, but the code I got from a class I had with one of the authors (authorization pending).

Below, I do my best to explain when/why use each of these and give examples of usage with the standard outputs from the sdm package, mainly from sdm() and predict().

## Installation

You can install the development version of sdmTools from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jfberner/sdmTools")
```

## Examples

### Base Objects

###### Disclaimer

- The following are the basic objects that are going to be used in all the examples. They are not meant to be fully reproducible, as species data, variable and algorithm selection vary considerably and should be picked according to each dataset and objectives. Instead, I intend to briefly take the reader through the process of creating an ENM with the sdm package solely for clarifying what each object is (or is supposed to be).

- Details on the nature and origin of the datasets should not be important for the use of this package's functions. 

- Mind that some lines of the code are a MUST if one's to implement my functions, and they are described as such.

The **'species'** I'm using here is going to be called 'charinus'. It is not a species, but is a model from another project I have easy at hand. I will not be sharing the dataset. I have two datasets: one called `occ_train` (with model training data) and another called `occ_test` (with model test data, which includes absence records). These objects are SpatialPointsDataFrame(s).

```{r echo=T, eval=T, results='hide'}
# The following are shapefiles (.shp) being loaded:
occ_train <- rgdal::readOGR('data/processed/shapefiles/train.shp')
occ_test <- rgdal::readOGR('data/processed/shapefiles/test.shp')
```

```{r eval=T}
# As SpatialPointsDataFrame
class(occ_train)
```

**Climate Data** (predictors) are four uncorrelated bioclimatic variables from [BioClim v2.1](https://www.worldclim.org). In the code they'll be called `envpres`, refferring to 'present' climate (average from 1970-2000, more on this on worldclim's website linked above).

```{r echo=T, eval=T, results='hide', warning=FALSE}
envpres <- dir(path = 'data/processed/envcropped/Present/', pattern = ".tif$", full.names = T)

envpres <- raster::stack(envpres)
```

```{r eval=T}
class(envpres)
```

I'll be using four **algorithms** in my all of the examples: SVM, MaxEnt, BioClim and Domain. References for those can be found in [e.g. @sobral-souzaSpeciesExtinctionRisk2015](https://doi.org/10.1016/j.ncon.2015.11.009), I will not go into details here.

```{r echo=T, results='hide'}
# Saving algorithm list can make life easier, for at least a couple reasons: 
# it retains the order in which they are modeled inside the sdm::sdm() function to be used in sdmTools functions, 
# and it saves you from re-typing them everytime.

algorithms <- c('svm', 'maxent', 'bioclim', 'domain') # order will be important
```

The **Model** will (naturally) be constructed using the sdm package. Details on sdm package usage can be found [here](https://www.youtube.com/watch?v=83dMS3bcjJM) in video format and [here](https://github.com/babaknaimi/sdm) is the github of the package. I won't go into further, unnecessary details.

First, we prepare the data:
```{r echo=T, results='hide'}
d_occ <- sdm::sdmData(formula = charinus~., train=occ_train,
                 predictors = envpres, bg = 18, # same number as occurrences
                 method = 'eRandom') # Create sdmData object
# eRandom = random in Environmental Space
```

Then we run the models:
```{r echo=T, results='hide'}
m_occ <- sdm::sdm(formula = charinus~.,data = d_occ,
              methods = algorithms,
                           replication = c('bootstrapping'), n=5)
```

This `m_occ` object is our sdmModel object, and it'll be often used.

Finally, we **predict** using built models:
```{r echo=T, results='hide'}
p_occ <- sdm::predict(m_occ, newdata = envpres, # new data is the environment in which we want to predict
                 filename = 'data/processed/model-build/predictions/predictions.present-rasterStd.tif',
                 prj = T, overwrite = T, nc = 7)

# This creates a file in the specified folder BUT the file doesn't retain the layer names, which will be super important for us to distinguish among methods. To do so:
```

This will save the predictions as a raster file. However, **layer names WILL NOT BE RETAINED**. Not only is retaining layer names extremely important to implement the functions of sdmTools package, it will make it easier for the user/researcher to distinguish them later. **To retain them**, simply:

```{r echo=T, results='hide'}
# The next line grabs the 'fullname' inside the p_occ object
names(p_occ) <- p_occ@z$fullname # This is SUPER important. It ensure both SPECIES and ALGORITHM names are retained in layer names

p_occ_stack <- raster::stack(p_occ) #convert rasterBrick to rasterStack
p_occ_spatraster <- terra::rast(p_occ_stack) #convert rasterStack to SpatRaster
names(p_occ_spatraster) <- names(p_occ_stack) #grab the names again, because in converting the terra::rast() function grabs the names from the .tif file that p_occ refers to.

terra::writeRaster(x = p_occ_spatraster, 
                    filename = 'data/processed/model-build/predictions/predictions.present-terraStd.tif',
                    overwrite = TRUE) # Using terra package because it retains layer names
```

When you read your predictions raster file, make sure to so using the terra package and then converting it to a RasterStack to retain sdm package standards (not required, but I haven't tested).

```{r echo=T, results='hide'}
pocc <- terra::rast(x = 'data/processed/model-build/predictions/predictions.present-terraStd.tif')
pocc <- raster::stack()
```

**Note:** there are now **two different prediction objects**, and they are **significantly different**. The object `p_occ` is the one originally obtained from the sdm::predict() function, and `pocc` is the one read from the terra package. Since is always best practice to have separate scripts for different steps in modelling **I'll be using `pocc` for all functions from here on**.

And these are all the objects we need to implement the functions of the sdmTools package: `occ_train`, `occ_test`, `envpres`, `m_occ` and `pocc`. The following sections will be examples of how to use the functions and a brief description, in my words, of when to use them.

```{r eval=T, echo=F,results=F}
pocc <- terra::rast(x = 'data/processed/model-build/predictions/predictions.present-all-algorithms.tif') # remember to read with terra:: to retain layer names
pocc <- raster::stack(pocc) # and then turn them into RasterStacks
m_occ <- sdm::read.sdm("data/processed/model-build/model-object/model.sdm")
```

### Boyce Index

The Boyce Index [@hirzelEvaluatingAbilityHabitat2006](https://doi.org/10.1016/j.ecolmodel.2006.05.017) DOES THIS AND THAT AND SHOULD BE USED FOR CASES WHERE X. DONE FOR TODAY.

Usage is as follows:

```{r example}
library(sdmTools)
boyce.output <- sdm_to_boyceIndex(preds = pocc, layer = mean(1:5), occurrence = occ_test)
# preds = pocc note that it is not p_occ, refer to the text above for the difference.
# layer = mean(1:5) will take the mean of the first five layers, in this case the five SVM models
```


# References
